Shapes analysis for time series.

### 2024-10-24 NEURIPS24/appendix.tex
add appendix

Given such a kernel $k:(\Rset^{d+1} \times \mathbb{S}^d)^2\to \Rset$
verifying \citep[Proposition 2 \& 4]{kaltenmark2017general}, we have that for any $(x,v)\in \Rset^{d+1}\times \mathbb{S}^{d}$, $\delta_{(x,\overrightarrow{v})}$ belongs to $\msw^*$ as a distribution and that the dual metric $\langle\cdot,\cdot \rangle_{\msw^*} $ satisfies for any $(x_1,v_1),(x_2,v_2)\in \left(\Rset^{d+1}\times \mathbb{S}^{d}\right)^2$,
\begin{equation}
  \langle\delta_{(x_1,\overrightarrow{v}_1)},\delta_{(x_2,\overrightarrow{v}_2)} \rangle_{\msw^*} =k((x_1,\overrightarrow{v}_1),(x_2,\overrightarrow{v}_2)) \eqsp .
\end{equation}
Thus, given two sets of triplets $X=(l_i,x_i,\overrightarrow{v}_i)_{i\in[T_0-1]}\in (\Rset\times \Rset^{d+1}\times \mathbb{S}^d)^{T_0-1},Y=(l_i',y_i,\overrightarrow{w}_i)_{i\in[T_1]}\in (\Rset\times \Rset^{d+1}\times \mathbb{S}^d)^{T_1-1}$ and denoting by 
\begin{equation}
  \label{eq:def_mu_X}
  \mu_X=\sum_{i=1}^{T_0} l_i \delta_{(x_i,\overrightarrow{v}_i)},\mu_Y=\sum_{i=1}^{T_1} l_i' \delta_{(y_i,\overrightarrow{w}_i)} \eqsp,
\end{equation}
 we have,
\begin{equation}
   |\mu_X-\mu_Y |_{\msw^*}^2 = \sum_{i,j = 1}^{T_0-1}l_i k((x_i,\overrightarrow{v_i}),(x_i,\overrightarrow{v_i}^0))l_j - 2 \sum_{i=1}^{T_0-1}\sum_{j=1}^{T_1-1}l_i k((x_i,\overrightarrow{v_i}),(y_i,\overrightarrow{w_i}))l_j' + \sum_{i,j = 1}^{T_1-1}l_i' k((y_i,\overrightarrow{w_i}),(y_i,\overrightarrow{w_i}))l_j' \eqsp.
\end{equation}
Then, using the identification $X\to \mu_X, Y\to \mu_Y$, we can define a distance on sets of triplets as $d_{\msw^*,3}(X,Y)=|\mu_X-\mu_Y|_{\msw^*}^2$.

@@

\subsection{Benchmark metohds}
\begin{itemize}
  \item SRV-based method: we include TCLR~\cite{heo2024logistic} a logistic regression on the tangent space of the Frechet mean with Square Root Velocity (SRV representation). We also include Shape-FPCA~\cite{wu2024shape} that encodes both the time series and its time parameterization. 
  \item LDDMM-Based : TS-LDDMM (ours) and LDDMM~\cite{glaunes2008large}. Both methods learn representations by solving ODEs parametrized with Kernels. While both methods handle multivariate signals, TS-LDDMM is specifically designed for time series.
 \end{itemize}

@@

\paragraph{Protocole}
For each dataset and method, the evaluation protocol is a simple train,validation test with hyperparameter tuning:
\begin{enumerate}
  \item Split The dataset in train 75\%, validation 15\%, and test 15\%.
  \item Training and hyperparameters tuning with train and validation sets
  \item Evaluate the macro f1-score on the test set 
\end{enumerate}

@@

@@

This experiment involves TS-LDDMM and LDDMM~\cite{glaunes2008large} methods. Both methods are run twice, first on respiratory cycles before exposure to the irritant molecule to capture mice breathing behavior at rest and on all respiratory cycles to capture the influence of the irritant molecule. Exposure to the irritant molecule leads to significant shape deformation in the respiratory cycles, and the terms must be added to the varifold loss to capture deformations at a large time scale.

@@

>>>>

Given such a kernel $k:(\Rset^{d+1} \times \mathbb{S}^d)^2\to \Rset$
verifying \citep[Proposition 2 \& 4]{kaltenmark2017general}, we have that for any $(x,v)\in \Rset^{d+1}\times \mathbb{S}^{d}$, $\delta_{(x,\overrightarrow{v})}$ belongs to $\msw^*$ as a distribution and that the dual metric $\langle\cdot,\cdot \rangle_{\msw^*} $ satisfies for any $(x_1,v_1),(x_2,v_2)\in \left(\Rset^{d+1}\times \mathbb{S}^{d}\right)^2$,
\begin{equation}
  \langle\delta_{(x_1,\overrightarrow{v}_1)},\delta_{(x_2,\overrightarrow{v}_2)} \rangle_{\msw^*} =k((x_1,\overrightarrow{v}_1),(x_2,\overrightarrow{v}_2)) \eqsp .
\end{equation}
Thus, given two sets of triplets $X=(l_i,x_i,\overrightarrow{v}_i)_{i\in[T_0-1]}\in (\Rset\times \Rset^{d+1}\times \mathbb{S}^d)^{T_0-1},Y=(l_i',y_i,\overrightarrow{w}_i)_{i\in[T_1]}\in (\Rset\times \Rset^{d+1}\times \mathbb{S}^d)^{T_1-1}$ and denoting by 
\begin{equation}
  \label{eq:def_mu_X}
  \mu_X=\sum_{i=1}^{T_0} l_i \delta_{(x_i,\overrightarrow{v}_i)},\mu_Y=\sum_{i=1}^{T_1} l_i' \delta_{(y_i,\overrightarrow{w}_i)} \eqsp,
\end{equation}
 we have,
\begin{equation}
  \begin{array}{ll}
    |\mu_X-\mu_Y |_{\msw^*}^2 = & \sum_{i,j = 1}^{T_0-1}l_i k((x_i,\overrightarrow{v_i}),(x_i,\overrightarrow{v_i}^0))l_j + \sum_{i,j = 1}^{T_1-1}l_i' k((y_i,\overrightarrow{w_i}),(y_i,\overrightarrow{w_i}))l_j' \\
    & - 2 \sum_{i=1}^{T_0-1}\sum_{j=1}^{T_1-1}l_i k((x_i,\overrightarrow{v_i}),(y_i,\overrightarrow{w_i}))l_j' \eqsp.
  \end{array}
\end{equation}
Then, using the identification $X\to \mu_X, Y\to \mu_Y$, we can define a distance on sets of triplets as $d_{\msw^*,3}(X,Y)=|\mu_X-\mu_Y|_{\msw^*}^2$.

@@

\subsection{Benchmark methods}
\begin{itemize}
  \item SRV-based method: we include TCLR~\cite{heo2024logistic} a logistic regression on the tangent space of the Frechet mean with Square Root Velocity (SRV representation). We also include Shape-FPCA~\cite{wu2024shape} that encodes both the time series and its time parameterization. 
  \item LDDMM-Based : TS-LDDMM (ours) and LDDMM~\cite{glaunes2008large}. Both methods learn representations by solving ODEs parametrized with Kernels. While both methods handle multivariate signals, TS-LDDMM is specifically designed for time series.
 \end{itemize}

@@

\subsection{Protocol}
For each dataset and method, the evaluation protocol is a simple train,validation test with hyperparameter tuning:
\begin{enumerate}
  \item Split The dataset in train 75\%, validation 15\%, and test 15\%.
  \item Training and hyperparameters tuning with train and validation sets
  \item Evaluate the macro f1-score on the test set 
\end{enumerate}

@@

\section{Appendix for the experiment: Noise sensitivity}
This experiment evaluates the influence of noise on the learning of the reference sequence for TS-LDDMM and SRVF Kacher-mean, a subroutine of Shape-FPCA~\cite{wu2024shape}.

\subsection{Protocol}
The dataset includes 100 sine waves with randomly generated time parametrization by following the procedure described in \Cref{appendix: settings_identifiability} with $\text{Gen}(50,1,0)$ and uniformly resampled. The dataset has been altered under four scenarios with an additive Gaussian noise centered and with standard deviation $\sigma_\epsilon \in \{0,0.05,0.1,0.2\}$. The referent sequence is learned for each scenario, and the $L_2$-norm error between the exact and the learned barycenter is computed.

\subsection{Method settings}
For SRVF, the number of steps to learn the Kacher-mean is set to 20. Regarding TS-LDDMM, the velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,65,1,1)$, and the varifold loss kernels $(k_{pos},k_{dir})$ are set to $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x}) = (5,1,1,1)$. For TS-LDDMM, the optimizer is set with 400 epochs and a maximum learning rate $\eta_M = 0.1$. In all cases, the initial reference graph is selected in the dataset as a time series with the median length.

\begin{figure}
  \centering
  \includegraphics[width = 0.9\linewidth]{pictures/noise sensitivity/noise sensitivity.pdf}
  \caption{Illustration of the learned barycenter (red) compared to the exact barycenter (green) for both TS-LDDMM \textbf{(a)} and Shape-FPCA \textbf{(b)}. The computation has been done for different level of noise $\epsilon \sim \mathcal{N}(0,\sigma_\epsilon)$ with $\sigma_\epsilon \in \{0,0.05,0.1,0.2\}$.}
  \label{fig: noise sensitivity}
\end{figure}

\subsection{Results}
\Cref{fig: noise sensitivity} illustrates the results. Noise level affects the learning reference graph in both cases, as depicted by the increasing error and the illustrations. However, the overall sine wave shape is better preserved by TS-LDDMM compared to SRVF Kacher-mean, for which the sine wave amplitude decreases as the noise increases. In addition, for TS-LDDMM, the regularity of the reference graph can be controlled by penalizing the norm of the velocity fields in the loss function. Further work on penalization will be conducted to handle noisy data better.

@@

This experiment involves TS-LDDMM, LDDMM~\cite{glaunes2008large} and Shape-FPCA~\cite{wu2024shape} methods. Two scenarios are investigated: before drug exposure and before/after drug exposure. All methods are investigated on both scenarios.

@@

\paragraph{Shape-FPCA parameters.}
For both scenarios, respiratory cycles are linearly interpolated and resampled to 200 points, and the length of the original time interval is kept. The computation of the Kacher-mean is done in a maximum of 50 iterations, and srv representations of the realigned time series and time parametrization are concatenated with cycle durations. When concatenating these vectors, the choice of amplitude factors is made to minimize the reconstruction error from the principal components analysis by following the procedure described in \cite{wu2024shape}. Shape-FPCA does not handle multivariate data, and we only kept the nasal airflow for this method.

\subsection{Addiotinal results}
\Cref{fig: after tslddmm} presents results for TS-LDDMM and \Cref{fig: after shapefpca}  presents results for Shape-FPCA.
The main components look similar. However, a subtle difference, yet important, can be noticed. With Shape-FPCA, the deformation tends to be a uniform time scaling, whereas, with TS-LDDMM, the time dilatation mainly occurs during the pause between inspiration and expiration. Qualitatively, this last deformation fits the physiological phenomenon: Mice's muscles cannot relax after exposure to the irritant molecule, leading to pauses between inspiration and expiration~\cite{nervo2019respiratory}. Qualitatively, contrary to Shape-FPCA, which manages to represent the main phenomena in the data, the deformations of TS-LDDMM capture subtle physiological behaviors essential for understanding the phenomenon at hand.

\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{pictures/exp after/exp after.pdf}
  \caption{Analysis of the first Principal Component (PC1) related to mice ventilation before and after exposure with TS-LDDMM representations. \textbf{(a)} displays PC densities per mice genotype, \textbf{(b)} illustrates deformations of the reference respiratory cycle $\mathbf{c}_0$ along PC1, and \textbf{(c)} displays all respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig: after tslddmm}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{pictures/exp after/exp after fpca.pdf}
  \caption{Analysis of the first Principal Component (PC1) related to mice ventilation before and after exposure with shape-FPCA representations. \textbf{(a)} displays PC densities per mice genotype, \textbf{(b)} illustrates deformations of the reference respiratory cycle $\mathbf{c}_0$ along PC1, and \textbf{(c)} displays all respiratory cycles with respect to time in PC1 and PC2 coordinates}
  \label{fig: after shapefpca}
\end{figure}

### 2024-10-24 NEURIPS24/experiment.tex
add appendix

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
    \caption{ColQ cycle}
    \label{fig:colq-cycle}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
    \caption{PC1 vs PC2}
    \label{fig: pcs-scatter}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
    \caption{WT cycle}
    \label{fig:wt-reference}
  \end{subfigure}
  \caption{\textbf{(a)} a ColQ respiratory cycle sample. \textbf{(b)} Referent respiratory cycle of individual mouse $\mathbf{c}_0^j$ in the TS-LDDMM PC1-PC2 coordinates system of $\mathbf{c}_0$. \textbf{(c)} a WT respiratory cycle sample.}
  \vspace{-1.5em}
\end{figure*}

\subsection{Interpretability: mice ventilation analysis}
\label{section:interpretability}

This experiment highlights the \textit{interpretability} of TS-LDDMM representation for studying the inter-individual variability in biomedical applications. We consider a time series dataset monitoring the evolution of mice's nasal and thoracic airflow when exposed to a drug altering respiration \cite{nervo2019respiratory}.
The dataset includes recordings of 7 control mice (WT) and 7 mutant mice (ColQ) with an enzyme deficiency. The enzyme is involved in the respiration regulation, and the drug inhibits its activity. For each mouse, airflows were monitored for 15 to 20 minutes before the drug exposure and then for 35 to 40 minutes. A complete description of the dataset is given in the \Cref{appendix:mouse_dataset}.

\paragraph{Experimental protocol.} 
We considered two experimental scenarios; the first focuses on mice ventilation before exposure to explore the inter-individual and genotype-specific variabilities. The second focuses on whole recordings to analyze the evolution of mice's ventilation after drug exposure. In both cases, the baseline protocol consists of first extracting $N$ respiratory cycles from the datasets with the procedure described in \cite{germain2023unsupervised}. Then, learning the referent respiratory cycle $\mathbf{c}_0$ and the representations  of respiratory cycles $(\bm{\alpha}_0^j)_{j \in [N]}$ by solving \eqref{eq:general_optimization_problem} using TS-LDDMM. $\bm{\alpha}_0^j$ being the momentum of the initial velocity field of the geodesic encodings the diffeomorphisms mapping $\mathbf{c}_0$ to the $j^{th}$ respiratory cycle. Finally, performing a Kernel-PCA on the initial velocity fields~\eqref{eq:def_v0} belonging to $\msv$ and encoded by the pairs $(\bm{\alpha}_0^j,\mathbf{c}_0)_{j \in [N]}$. The first experiment includes $N_1 = 700$ cycles collected before exposure. The second experiment includes $N_2 = 1400$ cycles with 25\% (resp. 75\%) before (resp. after) exposure. We also performed the first experimental scenario with LDDMM representation, and \Cref{appendix: mice_exp_setting} describes the settings of both methods. Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales.

\paragraph{Geodesic shooting along principal component directions.}
Any principal component (PC), noted $v_0^{pc}$, from a kernel-PCA in $\msv$, is itself an initial velocity field encoded by a pair $(\mathbf{c}_0, \bm{\alpha}_0^{pc})$. PCs encode the principal axis of deformations, and it is possible to shoot along the geodesic they encode with the differential equations \eqref{eq:integration}, enabling interpretation of the main sources of deformations.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{pictures/exp after/exp after.pdf}
  \caption{Analysis of the first Principal Component (PC1) related to mice ventilation before and after exposure with TS-LDDMM representations. \textbf{(a)} displays PC densities per mice genotype, \textbf{(b)} illustrates deformations of the reference respiratory cycle $\mathbf{c}_0$ along PC1, and \textbf{(c)} displays all respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCA}
  \vspace{-1.5em}
\end{figure*}

\paragraph{Mice ventilation evolution after drug exposure.} 
This experiment focuses on the first principal components learned from TS-LDDDM representations of respiratory cycles randomly sampled before and after drug exposure. \Cref{fig:exp_2_PCA}a illustrates the geodesic shootings along PC1. Again, PC1 accounts for variations in respiratory cycle duration, but more importantly, it can be observed on the deformation at -1.5 $\sigma_{\text{PC}}$ the apparition of a long pause after inspiration. Congruently, \Cref{fig:exp_2_PCA}c indicates that pauses appear after drug exposure as cycles with negative PC1 values mainly occur after 20 minutes and present more variability along PC3. In addition, \Cref{fig:exp_2_PCA}b shows a bimodal distribution for WT mice with one of the peaks in the negative values. This peak was not observed in the previous experiment \Cref{fig:exp1}a. It indicates that pauses after inspiration are prevalent in WT mice after drug exposure. On the other hand, the distributions of ColQ mice's respiratory cycles along PC1 in both experiments are similar and account for the same deformation, suggesting that ColQ mice weakly react to the drug exposure as they already adapt their enzyme deficiency.

>>>>

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
    \caption{ColQ cycle}
    \label{fig:colq-cycle}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
    \caption{PC1 vs PC2}
    \label{fig: pcs-scatter}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
    \caption{WT cycle}
    \label{fig:wt-reference}
  \end{subfigure}
  \caption{\textbf{(a)} a ColQ respiratory cycle sample. \textbf{(b)} Referent respiratory cycle of individual mouse $\mathbf{c}_0^j$ in the TS-LDDMM PC1-PC2 coordinates system of $\mathbf{c}_0$. \textbf{(c)} a WT respiratory cycle sample.}
  \vspace{-1.5em}
\end{figure*}
\vspace{-1ex}
\subsection{Interpretability: mice ventilation analysis}
\label{section:interpretability}
\vspace{-1ex}
This experiment highlights the \textit{interpretability} of TS-LDDMM representation for studying the inter-individual variability in biomedical applications. We consider a time series dataset monitoring the evolution of mice's nasal and thoracic airflow when exposed to a drug altering respiration \cite{nervo2019respiratory}.
The dataset includes recordings of 7 control mice (WT) and 7 mutant mice (ColQ) with an enzyme deficiency. The enzyme is involved in the respiration regulation, and the drug inhibits its activity. For each mouse, airflows were monitored for 15 to 20 minutes before the drug exposure and then for 35 to 40 minutes. A complete description of the dataset is given in the \Cref{appendix:mouse_dataset}.
\vspace{-1ex}

\paragraph{Experimental protocol.} 
We considered two experimental scenarios; the first focuses on mice ventilation before exposure to explore the inter-individual and genotype-specific variabilities. The second focuses on whole recordings to analyze the evolution of mice's ventilation after drug exposure. In both cases, the baseline protocol consists of first extracting $N$ respiratory cycles from the datasets with the procedure described in \cite{germain2023unsupervised}. Then, learning the referent respiratory cycle $\mathbf{c}_0$ and the representations  of respiratory cycles $(\bm{\alpha}_0^j)_{j \in [N]}$ by solving \eqref{eq:general_optimization_problem} using TS-LDDMM. $\bm{\alpha}_0^j$ being the momentum of the initial velocity field of the geodesic encodings the diffeomorphisms mapping $\mathbf{c}_0$ to the $j^{th}$ respiratory cycle. Finally, performing a Kernel-PCA on the initial velocity fields~\eqref{eq:def_v0} belonging to $\msv$ and encoded by the pairs $(\bm{\alpha}_0^j,\mathbf{c}_0)_{j \in [N]}$. The first experiment includes $N_1 = 700$ cycles collected before exposure. The second experiment includes $N_2 = 1400$ cycles with 25\% (resp. 75\%) before (resp. after) exposure. We also performed the first experimental scenario with LDDMM representation, and \Cref{appendix: mice_exp_setting} describes the settings of both methods. Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales. in addition, In addition, \Cref{appendix: mice_exp_setting}  presents a comparison between TS-LDDMM and Shape-FPCA on the second scenario. 
\vspace{-1ex}

\paragraph{Geodesic shooting along principal component directions.}
Any principal component (PC), noted $v_0^{pc}$, from a kernel-PCA in $\msv$, is itself an initial velocity field encoded by a pair $(\mathbf{c}_0, \bm{\alpha}_0^{pc})$. PCs encode the principal axis of deformations, and it is possible to shoot along the geodesic they encode with the differential equations \eqref{eq:integration}, enabling interpretation of the main sources of deformations.
\vspace{-1ex}

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{pictures/exp after/exp after.pdf}
  \caption{Analysis of the first Principal Component (PC1) related to mice ventilation before and after exposure with TS-LDDMM representations. \textbf{(a)} displays PC densities per mice genotype, \textbf{(b)} illustrates deformations of the reference respiratory cycle $\mathbf{c}_0$ along PC1, and \textbf{(c)} displays all respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCA}
  \vspace{-1.5em}
\end{figure*}
\vspace{-1ex}

\paragraph{Mice ventilation evolution after drug exposure.} 
This experiment focuses on the first principal components learned from TS-LDDDM representations of respiratory cycles randomly sampled before and after drug exposure. \Cref{fig:exp_2_PCA}a illustrates the geodesic shootings along PC1. Again, PC1 accounts for variations in respiratory cycle duration, but more importantly, it can be observed on the deformation at -1.5 $\sigma_{\text{PC}}$ the apparition of a long pause after inspiration. Congruently, \Cref{fig:exp_2_PCA}c indicates that pauses appear after drug exposure as cycles with negative PC1 values mainly occur after 20 minutes and present more variability along PC3. In addition, \Cref{fig:exp_2_PCA}b shows a bimodal distribution for WT mice with one of the peaks in the negative values. This peak was not observed in the previous experiment \Cref{fig:exp1}a. It indicates that pauses after inspiration are prevalent in WT mice after drug exposure. On the other hand, the distributions of ColQ mice's respiratory cycles along PC1 in both experiments are similar and account for the same deformation, suggesting that ColQ mice weakly react to the drug exposure as they already adapt their enzyme deficiency. 
\vspace{-1ex}
